For Finitary Induction-Induction, \\ Induction is Enough

### 2020-04-29 paper/paper.tex
stuff

\section{The Theory of IIT Signatures}
\label{sec:theory_of_signatures}

TODO: rewrite.

In this section we show how to describe signatures for IITs using a
QIIT, the theory of IIT signatures. We will say what it means for a
model of ETT to support the theory of IIT signatures. First we need to
say what a \emph{theory of IIT signature algebra} is and what
morphisms between such algebras are. Then supporting the theory of IIT
signatures means that there exists an algebra with a unique morphism
to every other algebra.

This description is analogous to saying that a model of ETT supports
natural numbers as follows: Using the internal language of the model,
a natural number algebra is given by a $\mathsf{Nat}:\Set$,
$\mathsf{zero}:\mathsf{Nat}$ and
$\mathsf{suc}:\mathsf{Nat}\ra \mathsf{Nat}$, while a morphism between
algebras $M$ and $N$ is given by a function
$f:\mathsf{Nat}^M\ra\mathsf{Nat}^N$ such that
$f\,\mathsf{zero}^M=\mathsf{zero}^N$ and
$f\,(\mathsf{suc}^M\,m)=\mathsf{suc}^N\,(f\,m)$. Natural numbers
themselves are given by an algebra equipped with a unique morphism
into any other algebra (the recursor).

We will call \emph{theory of IIT signature algebras} simply
\emph{signature algebras}. A signature algebra can be seen as a model
of a small type theory which consists of (1) a substitution calculus
(CwF \cite{Dybjer96internaltype}) equipped with (2) a universe, (3) a
function space where the domain is small and (3) another function
space with external domain. We explain the usage of these type
formers through examples after the definitions.

@@

\begin{definition}[IIT signature]
  An IIT signature is a context in an arbitrary signature algebra,
  that is, a function $(M:\SignAlg)\ra\Con^M$.
\end{definition}
Note that the notion of IIT signature is relative to a model of ETT:
it is expressed as a term (of a function type) in the model. This is
necessary because of the function space $\Pim$ which has as domain an
arbitrary type in the model. We make use of $\Pim$ in signatures with
external parameters, like the type of the elements in lists.

\begin{example}[Example signatures]\label{ex:signatures}
  The signature for natural numbers is given by the function
  \[
    \lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M).
  \]
  For readability we write the same signature without the $\lambda$
  and the superscripts below on the left and we also write an informal
  version with variable names on the right for comparison.
  \[
  \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
  \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
  \]
  This signature is given by a context of length three. We start with
  the empty context $\boldsymbol{\cdot}$, then we declare a sort by
  $\U$, then we declare an operator producing an element of $\El\,0$
  where $0$ is a de Bruijn index referring to the sort $\U$. Finally
  we declare an operator which takes as input an element of the sort
  $1$ and produces an element of the same sort. Note the assymetry of
  the function type $\Ra$: on the left there needs to be an element of
  $\U$, while on the right any type can appear (including another
  function type). This ensures strict positivity of the operators.

@@

\begin{example}[Algebras, displayed algebras and sections for natural
  numbers]
  For the signature of natural numbers as given in
  \Cref{ex:signatures}, algebras are given by the $\Sigma$-type
  $(N:\Set)\times N\times(N\ra N)$. A displayed algebra over $(N,z,s)$
  is given by the $\Sigma$-type
  \[
    (N^D:N\ra\Set)\times N^D\,z \times((n:N)\ra N^D\,n\ra N^D\,(s\,n)),
  \]
  and a section of a displayed algebra $(N^D,z^D,s^D)$ over $(N,z,s)$
  is given by the $\Sigma$-type
  \[
    (N^S:(n:N)\ra N^D\,n)\times (N^S\,z = z^D) \times((n:N)\ra N^S\,(s\,n) = s^D\,n\,(N^S\,n)).
  \]
  Displayed algebras over the initial algebra are called motives and
  methods of the eliminator, while a section of a displayed algebra
  over the initial algebra is the eliminator together with its
  computation rules.
\end{example}

@@

We first explain it by columns: the first column ($^A$ components)
corresponds to the standard model (set model, metacircular
interpretation \cite{ttintt}): contexts are sets, types are families,
terms are functions, the universe $\U$ is given by $\Set$, function
spaces are given by the external function space. The $^D$ column is a
logical predicate interpretation, $^\A$ and $^\D$ together are a unary
version of the parametric model for dependent types
\cite{10.1145/2535838.2535852}. Contexts are predicates, types are
families of predicates, terms say that the $^\A$ interpretation
respects the predicates (this is ususally called fundamental lemma of
the logical predicate). $\U$ is given by predicate space, the
predicate at a $\Pi$ type holds for a function if it respects the
predicates. For $\Pim$, the predicate is defined pointwise. The last
column ($^\S$) is a modified dependent logical relation which refers
to both $^A$ and $^\D$. Contexts are binary relations where the second
parameter depends on the first one, types are dependent variants of
this, terms say that the relation is respected by $^\A$ and $^\D$,
respectively. $\U$ is however not relation space, but a function and
$(\El\,a)^\S$ is the graph of the function $a^\S$. $\Pi^\S$ for a
function again says that the function respects the relation, however
we don't simply say
\[
  (\Pi\,a\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:a^\A\,\gamma)(\alpha^D:a^\D\,\gamma^D\,\alpha)(\alpha^\S:(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D)\ra B^\S\,\dots,
\]
as $(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D$ is just an equality
$a^\S\,\gamma^S\,\alpha = \alpha^D$ which we can singleton
contract. So we omit $\alpha^D$ and this equality as an input and
replace $\alpha^D$ by $a^\S\,\gamma^S\,\alpha$ in the definition.

@@

\begin{definition}[A model of ETT supports IITs]
  A model of ETT supports IITs if for any signature signature
  $\Omega:(M:\SignAlg)\ra\Con^M$ we have an $$
\end{definition}

Now, given a signature $f:(M:\SignAlg)\ra\Con^M$, we obtain
$(f\,\ADS)^\A$

  We describe the most important
  components, a full definition is given in the appendix (TODO). A
  $\Gamma:\Con$ is given by a $\Sigma$ type with three elements:
  \[
    (\Gamma^\A:\Set) \times (\Gamma^\D:\Gamma^\A\ra\Set) \times (\Gamma^\S:(\gamma:\Gamma^\A)\ra\Gamma^\D\,\gamma\ra\Set)
  \]
  As a signature is a context in an arbitrary signature algebra, using
  $\ADS$ a signature provides an algebra $\Gamma^\A$, a displayed
  algebra $\Gamma^\D$ and a section $\Gamma^\S$.

  A type in $\ADS$ is given by dependent versions of $\Con$:
  \begin{alignat*}{6}
    & (A:\Ty\,\Gamma) := \,\,&& (A^\A:\Gamma^A\ra\Set) \times (A^\D:\Gamma^\D\,\gamma\ra A^\A\,\gamma\ra\Set) \times \\
    & && (A^\S:\Gamma^\S\,\gamma\,\gamma^D\ra(\alpha:A^\A\,\gamma)\ra A^\D\,\gamma^D\,\alpha\ra\Set)
  \end{alignat*}
  Substitutions (and terms) are given by (dependent) functions which
  respect the predicates $^\D$ and relations $^\S$.
  \begin{alignat*}{10}
    & (\sigma:\Sub\,\Gamma\,\Delta) && := \,\, && (\sigma^\A:\Gamma^\A\ra\Delta^\A) \times (\sigma^\D:\Gamma^\D\,\gamma\ra\Delta^\D\,(\sigma^\A\,\gamma)) \times \\
    & && && \rlap{$(\sigma^\S:\Gamma^\S\,\gamma\,\gamma^D\ra\Delta^\S\,(\sigma^A\,\gamma)\,(\sigma^\D\,\gamma^D))$} \\
    & (t:\Tm\,\Gamma\,A) && := && (t^\A:(\gamma:\Gamma^\A)\ra A^\A\,\gamma) \times (t^\D:(\gamma^D:\Gamma^\D\,\gamma)\ra A^\D\,\gamma^D\,(t^\A\,\gamma)) \times \\
    & && && (t^\S:(\gamma^S:\Gamma^\S\,\gamma\,\gamma^D)\ra A^\S\,(t^A\,\gamma)\,(t^\D\,\gamma^D))
  \end{alignat*}

  xxx

>>>>

\section{A Definition of Inductive-Inductive Types}
\label{sec:theory_of_signatures}

In this section we say what it means that a model of ETT supports
IITs. We first define the notion of IIT signature. Signatures for
algebraic theories are usually given by inductive definitions. On the
one hand, we take this even further: our notion of signature is given
by a small type theory tailer-made to describe signatures, we call
this the \emph{theory of IIT signatures}. On the other hand we would
like to avoid using a complicated inductive definition (a type theory
is a quotient inductive-inductive type) to describe a simpler class of
inductive types. Hence we use a Church encoding of the theory of IIT
signatures to define signatures thus avoiding the need for
pre-existing inductive definitions. Another feature of our signatures
is that as they can include types inside the model of ETT (such as
$\N$ in the $\mathsf{isEven}$--$\mathsf{isOdd}$), hence signatures
arex internal to the particular model of ETT.

We define the theory of IIT signatures by saying what its algebras
(models) are. We call the \emph{theory of IIT signatures algebras}
simply \emph{signature algebras}. The theory of signatures is a small
type theory consisting of a (1) a substitution calculus (category with
families, CwF \cite{Dybjer96internaltype}) equipped with (2) a
universe, (3) a function space where the domain is in the universe and
(4) another function space with external domain. We explain the usage
of these type formers through examples after the definitions.

@@

\begin{definition}[IIT signature]
  An IIT signature is a context in an arbitrary signature algebra,
  that is, a function $(M:\SignAlg)\ra\Con^M$.
\end{definition}
This can be seen as a Church encoding of $\Con$. For our purposes it
is not a problem that signatures don't live at the lowest universe
level.

Note that the notion of IIT signature is relative to a model of ETT:
it is expressed as a term (of a function type) in the model. This is
necessary because of the function space $\Pim$ which has as domain an
arbitrary type in the model. We make use of $\Pim$ in signatures with
external parameters, like the type of the elements in lists.

\begin{example}[Example signatures]\label{ex:signatures}
  The signature for natural numbers is given by the function
  \[
    nat := \lambda M.(\boldsymbol{\cdot}^M\,\ext^M\,\U^M\,\ext^M\,\El^M\,0^M\,\ext^M\,1^M \Ra^M \El^M\,1^M).
  \]
  For readability we write the same signature without the $\lambda$
  and the superscripts below on the left and we also write an informal
  version with variable names on the right for comparison.
  \[
  \boldsymbol{\cdot}\,\ext\,\U\,\ext\,\El\,0\,\ext\,1 \Ra \El\,1 \hspace{5.9em}
  \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \Ra \El\,N
  \]
  This signature is given by a context of length three. We start with
  the empty context $\boldsymbol{\cdot}$, then we declare a sort by
  $\U$, then we declare an operator producing an element of $\El\,0$
  where $0$ is a de Bruijn index referring to the sort $\U$. Finally
  we declare an operator which takes as input an element of the sort
  $1$ and produces an element of the same sort. Note the assymetry of
  the function type $\Ra$: on the left there needs to be an element of
  $\U$, while on the right any type can appear (including another
  function type). This ensures strict positivity of the operators.

@@

\begin{example}[Algebras, displayed algebras and sections for natural
  numbers]\label{ex:adsnat}
  For the signature of natural numbers as given in
  \Cref{ex:signatures}, algebras are given by the $\Sigma$-type
  $(N:\Set)\times N\times(N\ra N)$. A displayed algebra over $(N,z,s)$
  is given by the $\Sigma$-type
  \[
    (N^D:N\ra\Set)\times N^D\,z \times((n:N)\ra N^D\,n\ra N^D\,(s\,n)),
  \]
  and a section of a displayed algebra $(N^D,z^D,s^D)$ over $(N,z,s)$
  is given by the $\Sigma$-type
  \[
    (N^S:(n:N)\ra N^D\,n)\times (N^S\,z = z^D) \times((n:N)\ra N^S\,(s\,n) = s^D\,n\,(N^S\,n)).
  \]
  Displayed algebras over the initial algebra are called motives and
  methods of the eliminator, while a section of a displayed algebra
  over the initial algebra is the eliminator together with its
  computation rules.
\end{example}

@@

We first explain it by columns: the first column ($^A$ components)
corresponds to the standard model (set model, metacircular
interpretation \cite{ttintt}): contexts are sets, types are families,
terms are functions, the universe $\U$ is given by $\Set$, function
spaces are given by the external function space. The $^\D$ column is a
logical predicate interpretation, $^\A$ and $^\D$ together are a unary
version of the parametric model for dependent types
\cite{10.1145/2535838.2535852}. Contexts are predicates, types are
families of predicates, terms say that the $^\A$ interpretation
respects the predicates (this is ususally called fundamental lemma of
the logical predicate). $\U$ is given by predicate space, the
predicate at a $\Pi$ type holds for a function if it respects the
predicates. For $\Pim$, the predicate is defined pointwise. The last
column ($^\S$) is a modified dependent logical relation which refers
to both $^A$ and $^\D$. Contexts are binary relations where the second
parameter depends on the first one, types are dependent variants of
this, terms say that the relation is respected by $^\A$ and $^\D$,
respectively. $\U$ is however not relation space, but a function and
$(\El\,a)^\S$ is the graph of the function $a^\S$. $\Pi^\S$ for a
function again says that the function respects the relation, however
we don't simply say
\[
  (\Pi\,a\,B)^\S\,\gamma^S\,f\,f^D := (\alpha:a^\A\,\gamma)(\alpha^D:a^\D\,\gamma^D\,\alpha)(\alpha^\S:(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D)\ra B^\S\,\dots,
\]
as $(\El\,a)^\S\,\gamma^S\,\alpha\,\alpha^D$ is just an equality
$a^\S\,\gamma^S\,\alpha = \alpha^D$ which we can singleton
contract. So we omit $\alpha^D$ and this equality as an input and
replace $\alpha^D$ by $a^\S\,\gamma^S\,\alpha$ in the definition.

@@

\begin{definition}[A model of ETT supports IITs]\label{sec:hasiits}
  A model of ETT supports IITs if for any signature
  $f:(M:\SignAlg)\ra\Con^M$ there is a $\con_f: (f\,\ADS)^\A$ and an
  $\elim_f : (\gamma^D:(f\,\ADS)^\D\,\con_f)\ra
  (f\,\ADS)^\S\,\con_f\,\gamma^D$.
\end{definition}
In other words, for any signature, we have an algebra called $\con$
(constructors) and for any displayed algebra over the constructors, we
have a section (called the eliminator).

One can check that \Cref{sec:hasiits} gives the right
notion of constructors and elimination principle for the signatures in
\Cref{ex:signatures}.
\begin{example}[An IIT supports natural numbers]
  For the signature $nat$ of natural numbers in \Cref{ex:signatures},
  the type of $\con_{nat}$ is
  \begin{alignat*}{5}
    & ({nat}\,\ADS)^\A = \\
    & (\boldsymbol{\cdot}^\ADS\,\ext^\ADS\,\U^\ADS\,\ext^\ADS\,\El^\ADS\,0^\ADS\,\ext^\ADS\,1^\ADS \Ra^\ADS \El^\ADS\,1^\ADS)^\A = \\
    & \Big(\big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\A = \\
    & \Big(\gamma'':\big(\gamma':((\gamma:\boldsymbol{\cdot}^\A)\times \U^\A\,\gamma)\big)\times (\El\,(\pi_2\,\id))^\A\,\gamma'\Big)\times\Big(\Pi\,\big(\pi_2\,(\pi_1\,\id)\big)\,\big(\pi_2\,(\pi_1\,(\pi_1\,\id))\big)\Big)^\A\,\gamma'' = \\
    & \Big(\gamma'':\big(\gamma':((\gamma:\top)\times\Set)\big)\times(\proj_2\,\gamma')\Big)\times\big(\proj_2\,(\proj_1\,\gamma'')\ra\proj_2\,(\proj_1\,\gamma'')\big),
  which is a left-nested $\Sigma$ type isomorphic to its
  right-nested counterpart
  \[
    (N:\Set)\times \big(N \times (N\ra N)\big).
  \]
  Writing $(((\tt,\mathsf{Nat}),\mathsf{zero}),\mathsf{suc})$ for
  $\con_{nat}$, the type of $\elim_{nat}$ computes as follows.
  \begin{alignat*}{5}
    & \rlap{$(\gamma^D: ({nat}\,\ADS)^\D\,\con_{nat})\ra ({nat}\,\ADS)^\S\,\con_{nat}\,\gamma^D =$} \\
    & \rlap{$\bigg(\gamma^D: \Big( \big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\D\,\con_{nat}\bigg)\ra$} \\
    & \rlap{$\Big( \big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\S\,\con_{nat}\,\gamma^D =$} \\
    & \rlap{$\bigg(\gamma^D: \Big( \big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\D\,$} \\
    & \rlap{$\hspace{1em} \big(((\tt,\mathsf{Nat}),\mathsf{zero}),\mathsf{suc}\big)\bigg)\ra$} \\
    & \rlap{$\Big(\big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\S\,\big(((\tt,\mathsf{Nat}),\mathsf{zero}),\mathsf{suc}\big)\,\gamma^D = $} \\
    & \bigg(\big(((\tt,N^D),z^D),s^D\big): && \Big( {\gamma^D}'':\big({\gamma^D}':((\gamma^D:\boldsymbol{\cdot}^\D\,\tt)\times \U^\D\,\gamma^D\,\mathsf{Nat})\big)\times (\El\,(\pi_2\,\id))^\D\,{\gamma^D}'\,\mathsf{zero}\Big)\times \\
    & && \Big(\Pi\,\big(\pi_2\,(\pi_1\,\id)\big)\,\big(\pi_2\,(\pi_1\,(\pi_1\,\id))\big)\Big)^\D\,{\gamma^D}''\,\mathsf{suc}\bigg)\ra \\
    & \rlap{$ \Big(\big((\boldsymbol{\cdot}\,\ext\,\U)\,\ext\,\El\,(\pi_2\,\id)\big)\,\ext\,\big(\pi_2\,(\pi_1\,\id)\big) \Ra \El\,\big(\pi_2\,(\pi_1\,\id)\big)\Big)^\S\,(((\tt,\mathsf{Nat}),\mathsf{zero}),\mathsf{suc})\,$} \\
    & \rlap{$\hspace{1em} \big(((\tt,N^D),z^D),s^D\big) =$} \\
    & \bigg(\big(((\tt,N^D),z^D),s^D\big): && \Big( {\gamma^D}'':\big({\gamma^D}':((\gamma^D:\boldsymbol{\cdot}^\D\,\tt)\times \U^\D\,\gamma^D\,\mathsf{Nat})\big)\times (\El\,(\pi_2\,\id))^\D\,{\gamma^D}'\,\mathsf{zero}\Big)\times \\
    & && \Big(\Pi\,\big(\pi_2\,(\pi_1\,\id)\big)\,\big(\pi_2\,(\pi_1\,(\pi_1\,\id))\big)\Big)^\D\,{\gamma^D}''\,\mathsf{suc}\bigg)\ra \\
    & \rlap{$ \Big( {\gamma^S}'':\big({\gamma^S}':((\gamma^S:\boldsymbol{\cdot}^\S\,\tt\,\tt)\times \U^\S\,\gamma^S\,\mathsf{Nat}\,N^D)\big)\times (\El\,(\pi_2\,\id))^\S\,{\gamma^S}'\,\mathsf{zero}\,z^D\Big)\times$} \\
    & \rlap{$ \Big(\Pi\,\big(\pi_2\,(\pi_1\,\id)\big)\,\big(\pi_2\,(\pi_1\,(\pi_1\,\id))\big)\Big)^\S\,{\gamma^S}''\,\mathsf{suc}\,s^D = $} \\
    & \bigg(\big(((\tt,N^D),z^D),s^D\big): && \Big( {\gamma^D}'':\big({\gamma^D}':((\gamma^D:\top)\times (\mathsf{Nat}\ra\Set))\big)\times \proj_2\,{\gamma^D}'\,\mathsf{zero}\Big)\times \\
    & && \big(\proj_2\,(\proj_1\,{\gamma^D}'')\,n\ra\proj_2\,(\proj_1\,{\gamma^D}'')\,(\mathsf{suc}\,n)\big)\bigg)\ra \\
    & \rlap{$ \Big( {\gamma^S}'':\big({\gamma^S}':((\gamma^S:\top)\times ((n:\mathsf{Nat})\ra N^D\,n))\big)\times \proj_2\,{\gamma^S}'\,\mathsf{zero} = z^D\Big)\times$} \\
    & \rlap{$ \Big(\big(n:\mathsf{Nat}\big)\ra\proj_2\,(\proj_1\,(\proj_1\,{\gamma^S}''))\,(\mathsf{suc}\,n) = s^D\,\big(\proj_2\,(\proj_1\,(\proj_1\,{\gamma^S}''))\,n\big)\Big)$}
  This is again a left-nested version of the expected elimination
  principle
    & (N^D:\mathsf{Nat}\ra\Set)(z^D:N^D\,\mathsf{zero})\big(s^D:(n:\mathsf{Nat})\ra N^D\,n\ra N^D\,(\mathsf{suc}\,n)\big)\ra \\
    & \big(N^S:(n:\mathsf{Nat})\ra N^D\,n\big)\times(N^S\,\mathsf{zero} = z^D)\times \big((n:\mathsf{Nat})\ra N^S\,(\mathsf{suc}\,n) = s^D\,(N^S\,n)\big)
\end{example}
\begin{remark}
  Note that as we work in a model of ETT, the computation rules that
  are expected from the elimination principle are given up to the
  internal equality type, however by equality reflection we also get
  them as definitional equalities.
\end{remark}
